---
title: "del7 final"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries}
library(rlang)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)
library(tibble)
library(pheatmap)
library(VennDiagram)
library(ggrepel)
library(ggpubr)
#library(Deducer)
library(broom)
library(boot)
library(data.table)
library(Rtsne)
library(multcomp)
library(FSA)
library(pvclust)
library(readxl)
library(stringr)
library(GGally)
library(xlsx)
library(readxl)
library(cluster)
library(snow)
library(UpSetR)
library(overLapPlot)
library(forcats)
#library(gridExtra) from upsetR
library(kableExtra)
```

##### 5mC analyses

Description of analysis (dataset is unpublished):

In brief, I had three meDIP-seq libraries to compare; a "wildtype" non-leukemic (wt), a "mutant" leukemic (R132H), and a novel "modified mutant" (del7) set. The "mutant" has been shown to be associated with a hypermethylation state at regulatory regions. The aim of the analysis was to examine the methylation state of this "modified mutant" relative to the "mutant".

To examine the differences in 5mC signal, I called differential methylation using the aforementioned pipeline between the wildtype and mutant. Using the hypermethylated regions between R132H vs wt as a reference, I measured the 5mC density at these same regions in the del7.

I then examined the same changes at key regulatory regions (hypermethylated CpG island promoters and active enhancers).

Lastly, I examined the changes in 5hmC at these same regions.

```{r 5mC analyses}
##### examining distribution of 5mC signal at hypermethylated regions
global_5mC <- read.delim("/home/jawong/src/del7/final/DMR/DMR_CpG/mIDH/hypermeth/DMR.union.RPKM", skip = 1, header = F) %>%
  setnames(c('chr',  'start', 'end' ,  'wt' ,   '72R132H'  , 'del7')) %>%
  gather(key = lib, value = rpkm, 4:9)

global_5mC %>%
  ggplot(., aes(x = lib, y = rpkm + 0.01)) +
  geom_boxplot() +
  coord_flip() +
  #scale_y_log10() +
  ylab("mC coverage (log rpkm)") + xlab("Libs (meDIP)") +
  theme_bw() + ylim(0, 500) +
  theme(text = element_text(size = 20))

##### 5mC signal at hypermethylated CpG islands and active enhancers ##### 

#enhancers
enh <- read.delim("/home/jawong/src/del7/final/DMR/DMR_CpG/mIDH/hypermeth/DMR.hyper_enhancers.RPKM", skip = 1, header = F) %>%
  setnames(c('chr',  'start', 'end' ,  'wt' ,     '72R132H'  , 'del7')) %>%
  gather(key = lib, value = rpkm, 4:9)

(gg_enh <- enh %>%
  ggplot(., aes(x = lib, y = rpkm + 0.01)) +
  geom_boxplot() +
  coord_flip() +
  scale_y_log10() +
  ylab("mC coverage (log rpkm)") + xlab("Libs (meDIP)") +
  theme_bw() +
  theme(text = element_text(size = 20)))

(gg_enh <- enh %>% dplyr::filter(lib %in% c("wt", "72R132H", "del7")) %>% 
                                   mutate(cell = ifelse(lib == "wt", "IDHwt",
                                                        ifelse(lib == "del7", "del7", "IDHm"))) %>%
    mutate(cell = factor(cell)) %>% 
    mutate(cell = factor(cell, levels(cell)[c(1,3,2)])) %>% 
  ggplot(., aes(x = cell, y = rpkm)) +
  geom_boxplot() +
  coord_flip() +
  ylab("5mC coverage (rpkm)") + xlab("Libs (meDIP)") +
  theme_bw() +
  theme(text = element_text(size = 20)))


cgip <- read.delim("/home/jawong/src/del7/final/DMR/DMR_CpG/mIDH/hypermeth/DMR.hyper_CGIpromoters.RPKM", skip = 1, header = F) %>%
  setnames(c('chr',  'start', 'end' ,  'wt' ,     '72R132H'  , 'del7')) %>%
  gather(key = lib, value = rpkm, 4:9)

(gg_cgip <- cgip %>%
  ggplot(., aes(x = lib, y = rpkm + 0.01)) +
  geom_boxplot() +
  coord_flip() +
  scale_y_log10() +
  ylab("mC coverage (log rpkm)") + xlab("Libs (meDIP)") +
  theme_bw() +
  theme(text = element_text(size = 20)))

(gg_cgip <- cgip %>% dplyr::filter(lib %in% c("wt", "72R132H", "del7")) %>% 
                                   mutate(cell = ifelse(lib == "wt", "IDHwt",
                                                        ifelse(lib == "del7", "del7", "IDHm"))) %>%
    mutate(cell = factor(cell)) %>% 
    mutate(cell = factor(cell, levels(cell)[c(1,3,2)])) %>% 
  ggplot(., aes(x = cell, y = rpkm)) +
  geom_boxplot() +
  coord_flip() +
  #scale_y_log10() +
  ylab("5mC coverage (rpkm)") + xlab("Libs (meDIP)") +
  theme_bw() +
  theme(text = element_text(size = 20)))

rbind(
  cgip %>% dplyr::filter(lib %in% c("wt", "72R132H", "del7")) %>% 
                                   mutate(cell = ifelse(lib == "wt", "IDHwt",
                                                        ifelse(lib == "del7", "del7", "IDHm"))) %>%
    mutate(cell = factor(cell)) %>% 
    mutate(cell = factor(cell, levels(cell)[c(1,3,2)])) %>%  mutate(region = "CGI promoter"),
  enh %>% dplyr::filter(lib %in% c("wt", "72R132H", "del7")) %>% 
                                   mutate(cell = ifelse(lib == "wt", "IDHwt",
                                                        ifelse(lib == "del7", "del7", "IDHm"))) %>%
    mutate(cell = factor(cell)) %>% 
    mutate(cell = factor(cell, levels(cell)[c(1,3,2)])) %>% mutate(region = "Enhancers")
) %>% 
  ggplot(., aes(x = cell, y = rpkm)) +
  geom_boxplot() +
  facet_wrap(~region, nrow = 2) +
  coord_flip() +
  #scale_y_log10() +
  ylab("5mC coverage (rpkm)") + xlab("Libs (meDIP)") +
  theme_linedraw() +
  theme(text = element_text(size = 20))

##### stat tests ######
#at enhancers
  enh %>% dplyr::filter(lib %in% c("wt", "72R132H", "del7")) %>% 
                                   mutate(cell = ifelse(lib == "wt", "IDHwt",
                                                        ifelse(lib == "del7", "del7", "IDHm"))) %>% 
  aov(data=., rpkm ~ lib) %>% 
  TukeyHSD()
#                    diff       lwr        upr    p adj
# del7-72R132H -51.819701 -55.23967 -48.399734 0.000000 ****
# wt-72R132H   -57.024197 -63.54778 -50.500617 0.000000 ****
# wt-del7       -5.204496 -11.76910   1.360113 0.150831

#at CGIs
  cgip %>% dplyr::filter(lib %in% c("wt", "72R132H", "del7")) %>% 
                                   mutate(cell = ifelse(lib == "wt", "IDHwt",
                                                        ifelse(lib == "del7", "del7", "IDHm"))) %>% 
  aov(data=., rpkm ~ lib) %>% 
  TukeyHSD()
#                    diff        lwr        upr    p adj
# del7-72R132H -64.096703 -66.630300 -61.563107 0.00e+00 ****
# wt-72R132H   -69.907681 -73.035056 -66.780305 0.00e+00 ****
# wt-del7       -5.810977  -8.980762  -2.641193 5.21e-05 *
```

The level of 5mC at enhancers and CGI promoters is significantly greater in R132H than in wt or del7. A small increase in 5mC levels in del7 relative to wt at CGI promoters was observed.

##### 5hmC analyses 

Examining 5hmC changes at these same hypermethylated regions. (Note I only had hmeDIP data for R132H and del7).

```{r 5hmc}
##### 5mC signal at hypermethylated regions ##### 
tmph <- read.delim("/home/jawong/src/del7/final/DMR/DMR_CpG/mIDH/hypermeth/DMR.union_hmc.RPKM", skip = 1, header = F) %>%
  setnames(c('chr',  'start', 'end' , 'IDHm unt', 'IDHm vitc', 'del7 unt', 'del7 vitc')) %>%
  gather(key = lib, value = rpkm, 4:7)

tmph %>%
  ggplot(., aes(x = lib, y = rpkm + 0.01)) +
  geom_boxplot() +
  coord_flip() +
  scale_y_log10() +
  ylab("hmC coverage (log rpkm)") + xlab("Libs (hmeDIP)") +
  theme_bw() +
  theme(text = element_text(size = 20))

enhh<- read.delim("/home/jawong/src/del7/final/DMR/DMR_CpG/mIDH/hypermeth/DMR.enh_hmc.RPKM", skip = 1, header = F) %>%
  setnames(c('chr',  'start', 'end' , 'IDHm unt', 'IDHm vitc', 'del7 unt', 'del7 vitc')) %>%
  gather(key = lib, value = rpkm, 4:7)

(gg_enhh <- enhh %>%
  ggplot(., aes(x = lib, y = rpkm + 0.01)) +
  geom_boxplot() +
  coord_flip() +
  scale_y_log10() +
  ylab("hmC coverage (log rpkm)") + xlab("Libs (hmeDIP)") +
  theme_bw() +
  theme(text = element_text(size = 20)))

cgiph <- read.delim("/home/jawong/src/del7/final/DMR/DMR_CpG/mIDH/hypermeth/DMR.cgip_hmc.RPKM", skip = 1, header = F) %>%
  setnames(c('chr',  'start', 'end' , 'IDHm unt', 'IDHm vitc', 'del7 unt', 'del7 vitc')) %>%
  gather(key = lib, value = rpkm, 4:7)

(gg_cgiph <- cgiph %>%
  ggplot(., aes(x = lib, y = rpkm + 0.01)) +
  geom_boxplot() +
  coord_flip() +
  scale_y_log10() +
  ylab("hmC coverage (log rpkm)") + xlab("Libs (hmeDIP)") +
  theme_bw() +
  theme(text = element_text(size = 20)))

rbind(
  cgiph %>% dplyr::filter(lib %in% c("IDHm unt", "del7 unt")) %>% 
                                   mutate(cell = ifelse(lib == "del7 unt", "del7", "IDHm")) %>%
    mutate(region = "CGI promoter"),
  enhh %>% dplyr::filter(lib %in% c("IDHm unt", "del7 unt")) %>% 
                                   mutate(cell = ifelse(lib == "del7 unt", "del7", "IDHm")) %>%
    mutate(region = "Enhancers")
) %>% 
  ggplot(., aes(x = cell, y = rpkm)) +
  geom_boxplot() +
  facet_wrap(~region, nrow = 2) +
  coord_flip() +
  #scale_y_log10() +
  ylab("5hmC coverage (rpkm)") + xlab("Libs (hmeDIP)") +
  theme_linedraw() +
  theme(text = element_text(size = 20))

###### 5hmC signal overall ####
#del7 vs 72hr R132H
rbind(
  read.delim("/home/jawong/src/del7/final/hmedip_DKO/union/DMR_fc/DHR/DMR.del7_untreated_hmedip.bwamem.sorted.dups_marked_72hr_R132H_unt_hmedip.del7_untreated_hmedip.bwamem.sorted.dups_marked.unique.bed", header = F) %>% 
  setnames(c('chr',  'start', 'end' , 'id', 'del7', 'IDHm', 'fc')) %>% 
  mutate(change = "lose 5hmc"),
   read.delim("/home/jawong/src/del7/final/hmedip_DKO/union/DMR_fc/DHR/DMR.del7_untreated_hmedip.bwamem.sorted.dups_marked_72hr_R132H_unt_hmedip.72hr_R132H_unt_hmedip.unique.bed", header = F) %>% 
    setnames(c('chr',  'start', 'end' , 'id', 'del7', 'IDHm', 'fc')) %>% 
  mutate(change = "gain 5hmc")
) %>% 
  gather(key = cell, value = rpkm, 5:6) %>% 
  ggplot(., aes(x = cell, y = rpkm, fill = change)) +
  #facet_wrap(~change) +
  geom_boxplot() + coord_flip() +
  scale_y_log10() + xlab("Libs (hmeDIP)") + ylab("5hmC coverage (log rpkm)") +
  theme_linedraw() +
  theme(text = element_text(size = 20))

#union file rpkm of peaks
read.delim("/home/jawong/src/del7/final/hmedip_DKO/union/DMR_fc/del7_untreated_hmedip.bwamem.sorted.dups_marked_72hr_R132H_unt_hmedip.union.RPKM", header = T) %>% 
  setnames(c('chr',  'start', 'end' , 'del7', 'IDHm')) %>% 
  gather(key = cell, value = rpkm, 4:5) %>% 
  ggplot(., aes(x = cell, y = rpkm + 0.01)) +
  geom_boxplot() + coord_flip() +
  scale_y_log10() + xlab("Libs (hmeDIP)") + ylab("5hmC coverage (log rpkm)") +
  theme_bw() +
  theme(text = element_text(size = 20), plot.margin = margin(10, 100, 10, 10))

t.test(read.delim("/home/jawong/src/del7/final/hmedip_DKO/union/DMR_fc/del7_untreated_hmedip.bwamem.sorted.dups_marked_72hr_R132H_unt_hmedip.union.RPKM", header = T) %>% setnames(c('chr',  'start', 'end' , 'del7', 'IDHm')) %>% 
         dplyr::select(del7) %>% unlist(),
       read.delim("/home/jawong/src/del7/final/hmedip_DKO/union/DMR_fc/del7_untreated_hmedip.bwamem.sorted.dups_marked_72hr_R132H_unt_hmedip.union.RPKM", header = T) %>% setnames(c('chr',  'start', 'end' , 'del7', 'IDHm')) %>% 
         dplyr::select(IDHm) %>% unlist(), 
       paired = T, alternative = "less")
# p < 2.2e-16

#del7 vs 15hr R132H
rbind(
  read.delim("/home/jawong/src/del7/final/hmedip_DKO/union/DMR_fc/DHR/DMR.del7_untreated_hmedip.bwamem.sorted.dups_marked_15hr_R132H_unt_hmedip.del7_untreated_hmedip.bwamem.sorted.dups_marked.unique.bed", header = F) %>% 
  setnames(c('chr',  'start', 'end' , 'id', 'del7', 'IDHm', 'fc')) %>% 
  mutate(change = "lose 5hmc"),
   read.delim("/home/jawong/src/del7/final/hmedip_DKO/union/DMR_fc/DHR/DMR.del7_untreated_hmedip.bwamem.sorted.dups_marked_15hr_R132H_unt_hmedip.15hr_R132H_unt_hmedip.unique.bed", header = F) %>% 
    setnames(c('chr',  'start', 'end' , 'id', 'del7', 'IDHm', 'fc')) %>% 
  mutate(change = "gain 5hmc")
) %>% 
  gather(key = cell, value = rpkm, 5:6) %>% 
  ggplot(., aes(x = cell, y = rpkm, fill = change)) +
  geom_boxplot() + coord_flip() +
  scale_y_log10() +
  theme_linedraw()
```

There is a significantly (p < 2.2e-16) lower level of 5hmC in del7 than in R132H.
